Error Reporting Parsers: a Monad Transformer Approach

### 2013-06-18 article.tex
Parsers:  explain how monad transformers help to build parser combinators

\section{Monad transformers}
Monad transformers~\cite{liang} provide a method of combining monads into
new monads which support operations of each of their constituents.  
By applying monad transformers, we can build semantically rich parsers without sacrificing modularity.
We'll use the standard transformers/mtl \cite{mtl} libraries -- 
they're easy to install, easy to use, and very useful.

First, note that the parser type we used earlier -- \verb+[t] -> Maybe ([t], a)+
-- can be built out of the StateT monad transformer applied to Maybe.
From the implementation:
\begin{verbatim}
newtype StateT s m a = StateT { runStateT :: s -> m (a,s) }
\end{verbatim}
Substituting in [s] for s and Maybe for m in the right-hand side produces
(note that the type variable m is no longer needed):
\begin{verbatim}
newtype StateT s a = StateT { runStateT :: [s] -> Maybe (a,[s]) }
\end{verbatim}
Which is identical to the Parser type given before, except for the order 
of the result tuple.
This allows us to define the new parser type as:
\begin{verbatim}
import Control.Monad.State       (StateT(..))

@@

item :: (MonadState [t] m, Alternative m) => m t
item =
    get >>= \xs -> case xs of
                        (t:ts) -> put ts *> pure t;
                        []     -> empty;
\end{verbatim}
Note that the FlexibleContexts language extension is required by the 
MonadState constraint in the type signature.  The mtl formulation of monad
transformers relies heavily on extensions to the type system, more of which 
we'll have to use for some of the later code in this article.

>>>>

\section{Monad transformers}

How can we use monad transformers to help us create parsers and parser combinators?
The parser type we used earlier -- \verb+[t] -> Maybe ([t], a)+
-- can be built out of the StateT monad transformer (from the standard 
mtl \cite{mtl} library) applied to Maybe.  Here's the StateT implementation:
\begin{verbatim}
newtype StateT s m a = StateT { runStateT :: s -> m (a,s) }
\end{verbatim}

Substituting in [s] for s and Maybe for m in the right-hand side produces
(note that the type variable m is no longer needed):
\begin{verbatim}
newtype StateT s a = StateT { runStateT :: [s] -> Maybe (a,[s]) }
\end{verbatim}

This is identical to the Parser type given before, except for the order 
of the result tuple, and so we can redefine the parser type as:
\begin{verbatim}
import Control.Monad.State       (StateT(..))

@@

item :: (MonadState [t] m, Alternative m) => m t
item =
    get >>= \xs -> case xs of
                        (t:ts) -> put ts *> pure t;
                        []     -> empty;
\end{verbatim}

There are two major benefits to our parsers from using monad transformers:
\begin{itemize}
 \item semantic building blocks -- the mtl library provides implementations
    for Maybe, Error, and State building blocks; using these, we can describe
    describe the semantics of our parsers
 \item type class instances -- in addition to providing the building blocks,
    the mtl provides instances for common type classes, so we only have to 
    worry about parser-specific operations
\end{itemize}
